Place of death in NRS COVID-19 deaths

### 2021-02-02 covid_nrs_place_of_death.Rmd
Reworking the script to reflect data availability from the two sources - doesn't run at the minute but just wanted to save changes before lunch

library(tidyverse)  # for tidy workflow
library(opendatascot)  # importing data from ScotGov open data website
library(phsmethods)  # methods for working with PHS data
# library(SPARQL)  # taken care of by opendatascot
library(lubridate)  # dealing with dates
library(janitor)  # for cleaning column names
library(ISOweek)  # for computing date from ISO week number + year
```

@@

# Import data

This analysis is based off two sets of data:

1. Weekly deaths involving COVID-19 from Scottish gov't open data service
  - this is imported using the `opendatascot` package in R, and connects to the data available at: [the Scottish Gov't open data   website](https://statistics.gov.scot/data/deaths-involving-coronavirus-covid-19)
2. Historical deaths 2015-2019, grouped by week & location
  - These are available at the NRS website on [deaths involving COVID-19]( https://www.nrscotland.gov.uk/statistics-and-data/statistics/statistics-by-theme/vital-events/general-publications/weekly-and-monthly-data-on-births-and-deaths/deaths-involving-coronavirus-covid-19-in-scotland), specifically under the 'Related Statistics' section, under the heading 'Ad-hoc queries'.
  - We load the total deaths, but also grouped by:
    - Health board
    - Local Authority
    - Sex
    - Age
    - ~~Ethnicity~~ Note: not sure this is available

Note that the weekly deaths due to COVID-19 also include a weekly average 2015-2019, but for the purposes of showing the range of deaths (weekly minima & maxima), we need the other datasets. The same holds for the age & sex breakdown.

@@

@@

# ods_structure("deaths-involving-coronavirus-covid-19")
# ods_schemes("deaths-involving-coronavirus-covid-19")
# ods_print_query("deaths-involving-coronavirus-covid-19")
ods_find_lower_geographies("S92000003")  # Scotland's refArea = S92000003

# TODO: split this into a scotland-wide dataset, and grouped by health board & council area
covid_deaths_raw <- bind_rows(
  ods_dataset("deaths-involving-coronavirus-covid-19",geography = "sc")#,
  # ods_dataset("deaths-involving-coronavirus-covid-19",geography = "la"),
  # ods_dataset("deaths-involving-coronavirus-covid-19",geography = "hb")
)
```

@@

```{r}
deaths_by_cause <- 
  covid_deaths_raw %>%
  clean_names() %>%
  rename(
    number_of_deaths = value,
    place_of_death = location_of_death  # so it's the same as everywhere else!
  ) %>%
  filter(!ref_period %in% as.character(2020:2021))  %>% # remove totals for 2020 & 2021
  mutate(
    year = parse_integer(str_extract(ref_period, pattern="^(2020|2021)")),
    week_number = parse_integer(str_remove(ref_period, pattern = "^(2020|2021)\\-")),
    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
    number_of_deaths = as.numeric(number_of_deaths)  # convert to numeric!
  ) %>%
  select(-c(ref_area, measure_type))  # ref_area is Scotland for all records; measure_type is count

## check date conversion
deaths_by_cause %>% count(ref_period, date_w_c) %>% slice(1:3, (nrow(.) - 3):nrow(.)) %>% knitr::kable(caption = "Checking the computation of 'date w/c' from year & week number - note how the 1st week of 2020 started in 2019.")
```

@@

```{r}
unique_places_of_death_from_data <- map(
  .x = list(deaths_by_cause, past_deaths_overall, past_deaths_health_board, past_deaths_council_area, past_deaths_age_sex),
  .f = function(x) {x$place_of_death}
  ) %>% unlist %>% unique %>% sort

@@

## this is dumb, but it works!
deaths_by_cause <- recode_place_of_death(deaths_by_cause)
past_deaths_overall <- recode_place_of_death(past_deaths_overall)
past_deaths_health_board <- recode_place_of_death(past_deaths_health_board)
past_deaths_council_area <- recode_place_of_death(past_deaths_council_area)
past_deaths_age_sex <- recode_place_of_death(past_deaths_age_sex)
past_deaths_overall_summarised <- recode_place_of_death(past_deaths_overall_summarised)

@@

```{r}
deaths_by_cause %>%
  filter(cause_of_death=="all-causes-average-of-corresponding-week-over-previous-5-years") %>%
  select(week_number, year = year, place_of_death_recoded, nrs_computed_mean = number_of_deaths) %>%
  left_join(past_deaths_overall_summarised) %>%
  filter(round(nrs_computed_mean) != round(deaths_mean))
  knitr::kable(caption = "Subset of weekly deaths where calculate mean deaths 2015-2019 differed between the COVID-19 related deaths file and the weekly deaths file. Week 53 was computed as the average of deaths in week 53 in 2015 and week 52 in 2016-2019. For all the other datasets the mean deaths computed agree!")
```

@@

merged_data_overall <-
  deaths_by_cause %>%
  split_deaths_due_to_covid_and_not %>%
  left_join(past_deaths_overall_summarised %>% mutate(sex = "all", age = "all"),  # only merge for aggregated data b/c that's all we have!
            by = c("place_of_death_recoded", "week_number", "sex", "age")) %>% 
  compute_run_over_week_for_2021

# TODO: get health board & council area in ODS export!
merged_deaths_by_health_board <-
  deaths_by_cause %>%
  split_deaths_due_to_covid_and_not %>%
  left_join(historical_range_by_health_board) %>%
  compute_run_over_week_for_2021

@@

removed_other_category_text <- glue::glue("Excluding 'Other' institutional place of death, N={merged_data %>% filter(sex == 'all' & age == 'all' & place_of_death_recoded == 'Other') %>% .$deaths_all_causes %>% sum} for weeks 1-{max(merged_data %>% filter(year==2020) %>% .$week_number)} in 2020 and 1-{max(merged_data %>% filter(year==2021) %>% .$week_number)} in 2021.")

@@

# figure_6_location_of_death <-
  merged_data %>%
  filter(sex == "all" & age == "all") %>%
  filter(!place_of_death_recoded %in% c("All","Other")) %>%
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid_related), 
               names_to = "deaths_type", values_to = "deaths") %>% 
  ggplot(aes(x = week_number_run_over)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  geom_line(aes(y = deaths_mean), linetype = "dashed", size = 0.2) +
  geom_vline(xintercept = 54, linetype = "dotted") +
  geom_text(data = annotations_figure_6, aes(x=x,y=y, label=label)) +
  scale_x_continuous(breaks = seq(0,60,by=10)) +
  facet_grid(~place_of_death_recoded) +
  theme_minimal() +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non Covid-19 deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
  labs(x = "Week number",
       y = "Deaths (N)",
       subtitle = date_range_text,
       caption = paste0(c(vertical_line_text, removed_other_category_text, source_nrs_text), collapse="\n"),
       colour = "Cause of death"
       )

@@

```{r}

>>>>

library(tidyverse)  # for tidy workflow
library(opendatascot)  # importing data from ScotGov open data website
library(phsmethods)  # methods for working with PHS data
library(readxl)  # for reading excel files
# library(SPARQL)  # taken care of by opendatascot
library(lubridate)  # dealing with dates
library(janitor)  # for cleaning column names
library(ISOweek)  # for computing date from ISO week number + year
```

@@

# Import & wrangle data

We're importing two kinds of data:

1. Weekly deaths involving COVID-19 from Scottish gov't open data service, broken down by week & place of death (home or institution)
2. Historical deaths 2015-2019, grouped by week & place of death

For both of these, we'll also look at crosstabulated figures for:

* Health board
* Local Authority
* Sex
* Age
* ~~Ethnicity~~ Note: not sure this is available

There are two ways of obtaining the data:

1. Using their SPARQL-based API ("data cube"), which can be accessed using the `opendatascot` package in R; the data can be seen on [the Scottish Gov't open data website](https://statistics.gov.scot/data/deaths-involving-coronavirus-covid-19)
2. There are a number of dataset, including the demographic/geographic crosstabulations available on the "Related Statistics" section of the [deaths involving COVID-19 website]( https://www.nrscotland.gov.uk/statistics-and-data/statistics/statistics-by-theme/vital-events/general-publications/weekly-and-monthly-data-on-births-and-deaths/deaths-involving-coronavirus-covid-19-in-scotland)

Note that the [*About* tab of the "data cube" interface](https://statistics.gov.scot/resource?uri=http%3A%2F%2Fstatistics.gov.scot%2Fdata%2Fdeaths-involving-coronavirus-covid-19) says the following about Health Board & Local Authority figures:

>Health Board and Local Authority figures include non-residents. Deaths are allocated to areas based on the usual residence of the deceased. If the deceased was not a Scottish resident, the death is allocated to the area where the death occurred.

This means we'll need to download those figures from the website rather than using the API

Note that the weekly deaths due to COVID-19 in 2020 & 2021 also include a weekly average 2015-2019, but for the purposes of showing the range of deaths (weekly minima & maxima), we need the weekly figures for individual years & the appropriate cross-tabulations.

## Parsing week numbers

In the `opendatascot`-imported data, `refPeriod` refers to the week number (called `week of occurrence` in the `.xlsx`/`.csv`-imported data has), as defined by [ISO 8601](https://en.wikipedia.org/wiki/ISO_week_date), and can be computed from a date with `lubridate::isoweek()`; the first week of 2020 is the week starting with Monday 2019-12-30, ~~and the corresponding start dates for each week can be computed by adding weeks using `lubridate::dweeks()` to that date.~~ This is non-trivial, because the number of iso weeks in a year isn't constant. The `ISOweek::ISOweek2date()` function deals with this issue in a crude manner, and allows us to compute the date (w/c on date) from the year & week number. It relies on being provided a valid week number (some years have 52 and some have 53), but it's probably safe to assume that the NRS data has valid dates.

We'll also compute a 'run-over' week number of 2021 in case we plot both years on a contibuous X-axis - in this case, week 1 of 2021 follows week 53 of 2020, so we start counting from 54 in 2021.

@@

~~Using `opendatascot` to load weekly deaths the level of geography can be specified, so we can have more or less granular data~~ Choosing a more granular level than Scotland-wide data means place of death is collapsed to the 'all' category only, see note above!

@@

# ods_structure("deaths-involving-coronavirus-covid-19")
# ods_schemes("deaths-involving-coronavirus-covid-19")
# ods_print_query("deaths-involving-coronavirus-covid-19")
# ods_print_query("deaths-involving-coronavirus-covid-19", geography = "la")
# ods_find_lower_geographies("S92000003")  # Scotland's refArea = S92000003
# phsmethods::match_area("S92000003")  # "Scotland"

make_human_readable_ref_area <- function(data_tbl, ref_area = refArea) {
  enquoted_ref_area <- enquo(ref_area)
  data_tbl %>%
    mutate(
      !!enquoted_ref_area := phsmethods::match_area(!!enquoted_ref_area)
    )
}
# raw_covid_deaths <- list(
#   scotland = ods_dataset("deaths-involving-coronavirus-covid-19", geography = "sc"),
#   health_board = ods_dataset("deaths-involving-coronavirus-covid-19", geography = "hb"),
#   local_authority = ods_dataset("deaths-involving-coronavirus-covid-19",geography = "la")
# ) %>%
#   map(.x = ., .f = ~make_human_readable_ref_area(data_tbl = .x))

weekly_deaths_by_age_sex <-
  ods_dataset("deaths-involving-coronavirus-covid-19", geography = "sc") %>%
  make_human_readable_ref_area %>%
  clean_names() %>%
  rename(
    number_of_deaths = value,
    place_of_death = location_of_death  # so it's the same as everywhere else!
  ) %>%
  filter(!ref_period %in% as.character(2020:2021)) %>% # remove totals for 2020 & 2021
  mutate(
    year = parse_integer(str_extract(ref_period, pattern="^(2020|2021)")),
    week_number = parse_integer(str_remove(ref_period, pattern = "^(2020|2021)\\-")),
    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
    number_of_deaths = as.numeric(number_of_deaths),  # convert to numeric!
    week_number_run_over = if_else(year == 2021, 53L + week_number, week_number)  # invent a "run-over" week number - weeks 54+ are in 2021
  ) %>%
  select(-c(measure_type))  #  measure_type is count for all data

weekly_deaths_by_la_file <-
  curl_download(url = "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-date-council-area-location.xlsx",
                destfile = tempfile(),
                quiet = FALSE)

weekly_deaths_by_la <- 
  read_excel(weekly_deaths_by_la_file, sheet = 2, skip = 2) %>% 
  clean_names %>%
  rename(week_number = week_of_occurrence, ref_area = council, place_of_death = location_of_death, number_of_deaths = deaths) %>%
  filter(!is.na(number_of_deaths)) %>%  # remove blank rows & copyright entry
  mutate(
    year = 2000L + parse_integer(str_sub(week_number, 1, 2)),
    week_number = parse_integer(str_sub(week_number, 4, 5)),
    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year)
  )

weekly_deaths_by_hb_file <-
  curl_download(url = "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-date-health-board-location.xlsx",
                destfile = tempfile(),
                quiet = FALSE)

weekly_deaths_by_hb <- 
  read_excel(weekly_deaths_by_hb_file, sheet = 2, skip = 2) %>% 
  clean_names %>%
  rename(week_number = week_of_occurrence, ref_area = health_board, place_of_death = location_of_death, number_of_deaths = deaths) %>%
  filter(!is.na(number_of_deaths)) %>%  # remove blank rows & copyright entry
  mutate(
    year = 2000L + parse_integer(str_sub(week_number, 1, 2)),
    week_number = parse_integer(str_sub(week_number, 4, 5)),
    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year)
  )

# TODO: split this into a scotland-wide dataset, and grouped by health board & council area
covid_deaths_raw <- bind_rows(
  ods_dataset("deaths-involving-coronavirus-covid-19",geography = "sc")#,
  # ods_dataset("deaths-involving-coronavirus-covid-19",geography = "la"),
  # ods_dataset("deaths-involving-coronavirus-covid-19",geography = "hb")
)

ods_dataset("deaths-involving-coronavirus-covid-19")

@@

```{r}
covid_deaths <-
  raw_covid_deaths %>%
  map(
    .x = .,
    .f = function(data_tbl) {
      data_tbl %>%
      clean_names() %>%
      rename(
        number_of_deaths = value,
        place_of_death = location_of_death  # so it's the same as everywhere else!
      ) %>%
      filter(!ref_period %in% as.character(2020:2021))  %>% # remove totals for 2020 & 2021
      mutate(
        year = parse_integer(str_extract(ref_period, pattern="^(2020|2021)")),
        week_number = parse_integer(str_remove(ref_period, pattern = "^(2020|2021)\\-")),
        date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
        number_of_deaths = as.numeric(number_of_deaths)  # convert to numeric!
      ) %>%
      select(-c(measure_type))  #  measure_type is count for all data
    }
  )

# TODO: remove this
deaths_by_cause <- 
  covid_deaths_raw %>%
  clean_names() %>%
  rename(
    number_of_deaths = value,
    place_of_death = location_of_death  # so it's the same as everywhere else!
  ) %>%
  filter(!ref_period %in% as.character(2020:2021))  %>% # remove totals for 2020 & 2021
  mutate(
    year = parse_integer(str_extract(ref_period, pattern="^(2020|2021)")),
    week_number = parse_integer(str_remove(ref_period, pattern = "^(2020|2021)\\-")),
    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
    number_of_deaths = as.numeric(number_of_deaths)  # convert to numeric!
  ) %>%
  select(-c(ref_area, measure_type))  # ref_area is Scotland for all records; measure_type is count

## check date conversion
covid_deaths$scotland %>% count(ref_period, date_w_c) %>% slice(1:3, (nrow(.) - 3):nrow(.)) %>% knitr::kable(caption = "Checking the computation of 'date w/c' from year & week number - note how the 1st week of 2020 started in 2019.")
```

@@

```{r}
unique_places_of_death_from_data <- map(
  .x = c(covid_deaths,list(past_deaths_overall, past_deaths_health_board, past_deaths_council_area, past_deaths_age_sex)),
  .f = function(x) {x$place_of_death}
  ) %>% unlist %>% unique %>% sort

@@

## this is dumb, but it works!
covid_deaths <- map(.x = covid_deaths, .f = ~recode_place_of_death(.x))
# deaths_by_cause <- recode_place_of_death(deaths_by_cause)
past_deaths_overall <- recode_place_of_death(past_deaths_overall)
past_deaths_health_board <- recode_place_of_death(past_deaths_health_board)
past_deaths_council_area <- recode_place_of_death(past_deaths_council_area)
past_deaths_age_sex <- recode_place_of_death(past_deaths_age_sex)
past_deaths_overall_summarised <- recode_place_of_death(past_deaths_overall_summarised)

@@

```{r}
covid_deaths$scotland %>%
  filter(cause_of_death=="all-causes-average-of-corresponding-week-over-previous-5-years") %>%
  select(week_number, year = year, place_of_death_recoded, nrs_computed_mean = number_of_deaths) %>%
  left_join(past_deaths_overall_summarised) %>%
  filter(round(nrs_computed_mean) != round(deaths_mean))
  knitr::kable(caption = "Subset of weekly deaths where calculate mean deaths 2015-2019 differed between the COVID-19 related deaths file and the weekly deaths file. Week 53 was computed as the average of deaths in week 53 in 2015 and week 52 in 2016-2019. For all the other datasets the mean deaths computed agree!")
```

@@

merged_data_overall <-
  covid_deaths$scotland %>%
  split_deaths_due_to_covid_and_not %>%
  left_join(past_deaths_overall_summarised %>% mutate(sex = "all", age = "all"),  # only merge for aggregated data b/c that's all we have!
            by = c("place_of_death_recoded", "week_number", "sex", "age")) %>% 
  compute_run_over_week_for_2021

# TODO: get health board & council area in ODS export!
merged_deaths_by_health_board <-
  covid_deaths$health_board %>%
  split_deaths_due_to_covid_and_not %>%
  # inner_join(historical_range_by_health_board, by = c("week_number","place_of_death_recoded","ref_area"="health_board")) %>%
  inner_join(historical_range_by_health_board, by = c("week_number","ref_area"="health_board","place_of_death_recoded")) %>%
  compute_run_over_week_for_2021

@@

removed_other_category_text <- glue::glue("Excluding 'Other' institutional place of death, N={merged_data_overall %>% filter(sex == 'all' & age == 'all' & place_of_death_recoded == 'Other') %>% .$deaths_all_causes %>% sum} for weeks 1-{max(merged_data_overall %>% filter(year==2020) %>% .$week_number)} in 2020 and 1-{max(merged_data_overall %>% filter(year==2021) %>% .$week_number)} in 2021.")

@@

# figure_6_location_of_death <-
  merged_data_overall %>%
  filter(sex == "all" & age == "all") %>%
  filter(!place_of_death_recoded %in% c("All","Other")) %>%
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid_related), 
               names_to = "deaths_type", values_to = "deaths") %>% 
  ggplot(aes(x = week_number_run_over)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  geom_line(aes(y = deaths_mean), linetype = "dashed", size = 0.2) +
  geom_vline(xintercept = 54, linetype = "dotted") +
  geom_text(data = annotations_figure_6, aes(x=x,y=y, label=label)) +
  scale_x_continuous(breaks = seq(0,60,by=10)) +
  facet_grid(~place_of_death_recoded) +
  theme_minimal() +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non Covid-19 deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
  labs(x = "Week number",
       y = "Deaths (N)",
       subtitle = date_range_text,
       caption = paste0(c(vertical_line_text, removed_other_category_text, source_nrs_text), collapse="\n"),
       colour = "Cause of death"
       )

@@

```{r}
merged_deaths_by_sex %>%
  filter(sex != "all" & age == "all") %>%
  filter(place_of_death_recoded == "Home & other non-institution") %>%
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid_related), 
               names_to = "deaths_type", values_to = "deaths") %>% 
  ggplot(aes(x = week_number_run_over)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  geom_line(aes(y = deaths_mean), linetype = "dashed", size = 0.2) +
  geom_vline(xintercept = 54, linetype = "dotted") +
  # geom_text(data = annotations_figure_6, aes(x=x,y=y, label=label)) +
  scale_x_continuous(breaks = seq(0,60,by=10)) +
  facet_wrap(~sex) +
  theme_minimal() +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non Covid-19 deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
  labs(x = "Week number",
       y = "Deaths (N)",
       subtitle = date_range_text,
       # caption = paste0(c(vertical_line_text, removed_other_category_text, source_nrs_text), collapse="\n"),
       colour = "Cause of death"
       )
```
